The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth

### 2021-09-22 paper/paper.tex

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (M. Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis (IDA)'' (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M. Huebner et al. 2020).

@@

Our process of cleaning builds heavily on the tidyverse approach (Wickham et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

This paper is structured in the following way. Section \ref{database} describes the NLSY79 data source. Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies. Our final subset is compared the old textbook subset in Section \ref{comare}. Finally, Section \ref{summary} summarises the contribution and makes recommendations for the NLSY79 data curators.

@@

\CommentTok{\#tidy the date of birth data}
\NormalTok{dob\_tidy }\OtherTok{\textless{}{-}}\NormalTok{ new\_data\_qnames }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(CASEID\_1979,}
                \FunctionTok{starts\_with}\NormalTok{(}\StringTok{"Q1{-}3\_A\textasciitilde{}"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{dob\_year =} \FunctionTok{case\_when}\NormalTok{(}
                      \CommentTok{\# if the years recorded in both sets match, take 79 data}
                      \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{==} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}}\NormalTok{,}
                      \CommentTok{\# if the year in the 81 set is missing, take the 79 data}
                      \FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}}\NormalTok{,}
                      \CommentTok{\# if the sets don\textquotesingle{}t match for dob year, take the 79 data}
                      \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}}\NormalTok{),}
         \AttributeTok{dob\_month =} \FunctionTok{case\_when}\NormalTok{(}
                      \CommentTok{\# if months recorded in both sets match, take 79 data}
                      \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{==} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}}\NormalTok{,}
                      \CommentTok{\# if month in 81 set is missing, take the 79 data}
                      \FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}}\NormalTok{,}
                      \CommentTok{\# if sets don\textquotesingle{}t match for dob month, take the 79 data}
                      \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}}\NormalTok{),}
         \CommentTok{\# flag if there is a conflict between dob recorded in 79 and 81}
         \AttributeTok{dob\_conflict =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{                      (}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\&} \SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}}\NormalTok{)}
                      \SpecialCharTok{\textasciitilde{}} \ConstantTok{TRUE}\NormalTok{,}
\NormalTok{                      (}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{!=} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\&} \SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}}\NormalTok{)}
                      \SpecialCharTok{\textasciitilde{}} \ConstantTok{TRUE}\NormalTok{,}
\NormalTok{                      (}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{==} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\&}
\NormalTok{                        (}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1979}\StringTok{\textasciigrave{}} \SpecialCharTok{==} \StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \ConstantTok{FALSE}\NormalTok{,}
                      \FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}M\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{|} \FunctionTok{is.na}\NormalTok{(}\StringTok{\textasciigrave{}}\AttributeTok{Q1{-}3\_A\textasciitilde{}Y\_1981}\StringTok{\textasciigrave{}}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \ConstantTok{FALSE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(CASEID\_1979,}
\NormalTok{         dob\_month,}
\NormalTok{         dob\_year,}
\NormalTok{         dob\_conflict)}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# make a list for years where we used the "QES{-}52A"}
\NormalTok{year\_A }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1979}\SpecialCharTok{:}\DecValTok{1987}\NormalTok{, }\DecValTok{1993}\NormalTok{)}
\CommentTok{\#function to get the hour of work}
\NormalTok{get\_hour }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(year)\{}
  \ControlFlowTok{if}\NormalTok{(year }\SpecialCharTok{\%in\%}\NormalTok{ year\_A)\{}
\NormalTok{    temp }\OtherTok{\textless{}{-}}\NormalTok{ new\_data\_qnames }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{select}\NormalTok{(CASEID\_1979,}
                    \FunctionTok{starts\_with}\NormalTok{(}\StringTok{"QES{-}52A"}\NormalTok{) }\SpecialCharTok{\&}
                      \FunctionTok{ends\_with}\NormalTok{(}\FunctionTok{as.character}\NormalTok{(year)))\}}
  \ControlFlowTok{else}\NormalTok{\{}
\NormalTok{    temp }\OtherTok{\textless{}{-}}\NormalTok{ new\_data\_qnames }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{select}\NormalTok{(CASEID\_1979,}
                    \FunctionTok{starts\_with}\NormalTok{(}\StringTok{"QES{-}52D"}\NormalTok{) }\SpecialCharTok{\&}
                      \FunctionTok{ends\_with}\NormalTok{(}\FunctionTok{as.character}\NormalTok{(year)))\}}
\NormalTok{  temp }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{pivot\_longer}\NormalTok{(}\SpecialCharTok{!}\NormalTok{CASEID\_1979,}
                 \AttributeTok{names\_to =} \StringTok{"job"}\NormalTok{,}
                 \AttributeTok{values\_to =} \StringTok{"hours\_work"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{separate}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\FunctionTok{c}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\StringTok{"year"}\NormalTok{), }\AttributeTok{sep =} \SpecialCharTok{{-}}\DecValTok{4}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{job =} \FunctionTok{paste0}\NormalTok{(}\StringTok{"job\_"}\NormalTok{, }\FunctionTok{substr}\NormalTok{(job, }\DecValTok{9}\NormalTok{, }\DecValTok{10}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{rename}\NormalTok{(}\AttributeTok{id =}\NormalTok{ CASEID\_1979)}
\NormalTok{\}}

\CommentTok{\# getting the hours of work of all observations}
\NormalTok{years }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1979}\SpecialCharTok{:}\DecValTok{1994}\NormalTok{, }\FunctionTok{seq}\NormalTok{(}\DecValTok{1996}\NormalTok{, }\DecValTok{2018}\NormalTok{, }\AttributeTok{by =} \DecValTok{2}\NormalTok{))}
\NormalTok{hours\_all }\OtherTok{\textless{}{-}} \FunctionTok{map\_dfr}\NormalTok{(years, get\_hour)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{get\_rate }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(year) \{}
\NormalTok{  new\_data\_qnames }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{select}\NormalTok{(CASEID\_1979,}
                  \FunctionTok{starts\_with}\NormalTok{(}\StringTok{"HRP"}\NormalTok{) }\SpecialCharTok{\&}
                    \FunctionTok{ends\_with}\NormalTok{(}\FunctionTok{as.character}\NormalTok{(year))) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{pivot\_longer}\NormalTok{(}\SpecialCharTok{!}\NormalTok{CASEID\_1979, }\AttributeTok{names\_to =} \StringTok{"job"}\NormalTok{, }\AttributeTok{values\_to =} \StringTok{"rate\_per\_hour"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{separate}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\FunctionTok{c}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\StringTok{"year"}\NormalTok{), }\AttributeTok{sep =} \SpecialCharTok{{-}}\DecValTok{4}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{job =} \FunctionTok{paste0}\NormalTok{(}\StringTok{"job\_0"}\NormalTok{, }\FunctionTok{substr}\NormalTok{(job, }\DecValTok{4}\NormalTok{, }\DecValTok{4}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{rename}\NormalTok{(}\AttributeTok{id =}\NormalTok{ CASEID\_1979)}
\NormalTok{\}}
\NormalTok{rates\_all }\OtherTok{\textless{}{-}} \FunctionTok{map\_dfr}\NormalTok{(years, get\_rate)}
\CommentTok{\# join hours and rates variable}
\NormalTok{hours\_wages }\OtherTok{\textless{}{-}} \FunctionTok{left\_join}\NormalTok{(rates\_all,}
\NormalTok{                         hours\_all,}
                         \AttributeTok{by =} \FunctionTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"year"}\NormalTok{, }\StringTok{"job"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
  \CommentTok{\# set the 0 value in rate\_per\_hour as NA}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{rate\_per\_hour =} \FunctionTok{ifelse}\NormalTok{(rate\_per\_hour }\SpecialCharTok{==} \DecValTok{0}\NormalTok{, }\ConstantTok{NA}\NormalTok{,}
\NormalTok{                                rate\_per\_hour),}
         \AttributeTok{hours\_work =} \FunctionTok{ifelse}\NormalTok{(hours\_work }\SpecialCharTok{==} \DecValTok{0} \SpecialCharTok{|}\NormalTok{ hours\_work }\SpecialCharTok{\textgreater{}} \DecValTok{84}\NormalTok{, }\ConstantTok{NA}\NormalTok{, hours\_work))}
\end{Highlighting}
\end{Shaded}

@@

{\centering \includegraphics[width=0.93\linewidth]{/Users/cookd/numbats/summer-wages-refresh/paper/figures/flowchart}

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. {``Robustlmm: An r Package for Robust Estimation of Linear Mixed-Effects Models.''} \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
Loo, Mark van der. 2018. \emph{Statistical Data Cleaning with Applications in r}.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render SVG Images into PDF, PNG, PostScript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.

@@

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. {``Robust Regression \textbar{} r Data Analysis Examples.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-validate}{}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

>>>>

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis (IDA)'' (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al. 2020).

@@

Our process of cleaning builds heavily on the tidyverse approach (H. Wickham, Averick, et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). The resulting data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

This paper is structured in the following way. Section \ref{database} describes the NLSY79 data source. Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies. Our final subset is compared the old textbook subset in Section \ref{comare}. Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.

@@

\CommentTok{#tidy the date of birth data}
\NormalTok{dob_tidy <-}\StringTok{ }\NormalTok{new_data_qnames }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(CASEID_}\DecValTok{1979}\NormalTok{,}
                \KeywordTok{starts_with}\NormalTok{(}\StringTok{"Q1-3_A~"}\NormalTok{)) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{dob_year =} \KeywordTok{case_when}\NormalTok{(}
                      \CommentTok{# if the years recorded in both sets match, take 79 data}
                      \StringTok{`}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`} \OperatorTok{==}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`} \OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`}\NormalTok{,}
                      \CommentTok{# if the year in the 81 set is missing, take the 79 data}
                      \KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`}\NormalTok{) }\OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`}\NormalTok{,}
                      \CommentTok{# if the sets don't match for dob year, take the 79 data}
                      \StringTok{`}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`} \OperatorTok{!=}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`} \OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`}\NormalTok{),}
         \DataTypeTok{dob_month =} \KeywordTok{case_when}\NormalTok{(}
                      \CommentTok{# if months recorded in both sets match, take 79 data}
                      \StringTok{`}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`} \OperatorTok{==}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`} \OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`}\NormalTok{,}
                      \CommentTok{# if month in 81 set is missing, take the 79 data}
                      \KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`}\NormalTok{) }\OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`}\NormalTok{,}
                      \CommentTok{# if sets don't match for dob month, take the 79 data}
                      \StringTok{`}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`} \OperatorTok{!=}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`} \OperatorTok{~}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`}\NormalTok{),}
         \CommentTok{# flag if there is a conflict between dob recorded in 79 and 81}
         \DataTypeTok{dob_conflict =} \KeywordTok{case_when}\NormalTok{(}
\NormalTok{                      (}\StringTok{`}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`} \OperatorTok{!=}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`}\NormalTok{) }\OperatorTok{&}\StringTok{ }\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`}\NormalTok{)}
                      \OperatorTok{~}\StringTok{ }\OtherTok{TRUE}\NormalTok{,}
\NormalTok{                      (}\StringTok{`}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`} \OperatorTok{!=}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`}\NormalTok{) }\OperatorTok{&}\StringTok{ }\OperatorTok{!}\KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`}\NormalTok{)}
                      \OperatorTok{~}\StringTok{ }\OtherTok{TRUE}\NormalTok{,}
\NormalTok{                      (}\StringTok{`}\DataTypeTok{Q1-3_A~Y_1979}\StringTok{`} \OperatorTok{==}\StringTok{ `}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`}\NormalTok{) }\OperatorTok{&}
\StringTok{                        }\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~M_1979}\StringTok{`} \OperatorTok{==}\StringTok{ `}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`}\NormalTok{) }\OperatorTok{~}\StringTok{ }\OtherTok{FALSE}\NormalTok{,}
                      \KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~M_1981}\StringTok{`}\NormalTok{) }\OperatorTok{|}\StringTok{ }\KeywordTok{is.na}\NormalTok{(}\StringTok{`}\DataTypeTok{Q1-3_A~Y_1981}\StringTok{`}\NormalTok{) }\OperatorTok{~}\StringTok{ }\OtherTok{FALSE}\NormalTok{)) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(CASEID_}\DecValTok{1979}\NormalTok{,}
\NormalTok{         dob_month,}
\NormalTok{         dob_year,}
\NormalTok{         dob_conflict)}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# make a list for years where we used the "QES-52A"}
\NormalTok{year_A <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1979}\OperatorTok{:}\DecValTok{1987}\NormalTok{, }\DecValTok{1993}\NormalTok{)}
\CommentTok{#function to get the hour of work}
\NormalTok{get_hour <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(year)\{}
  \ControlFlowTok{if}\NormalTok{(year }\OperatorTok{%in%}\StringTok{ }\NormalTok{year_A)\{}
\NormalTok{    temp <-}\StringTok{ }\NormalTok{new_data_qnames }\OperatorTok{%>%}
\StringTok{      }\KeywordTok{select}\NormalTok{(CASEID_}\DecValTok{1979}\NormalTok{,}
                    \KeywordTok{starts_with}\NormalTok{(}\StringTok{"QES-52A"}\NormalTok{) }\OperatorTok{&}
\StringTok{                      }\KeywordTok{ends_with}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(year)))\}}
  \ControlFlowTok{else}\NormalTok{\{}
\NormalTok{    temp <-}\StringTok{ }\NormalTok{new_data_qnames }\OperatorTok{%>%}
\StringTok{      }\KeywordTok{select}\NormalTok{(CASEID_}\DecValTok{1979}\NormalTok{,}
                    \KeywordTok{starts_with}\NormalTok{(}\StringTok{"QES-52D"}\NormalTok{) }\OperatorTok{&}
\StringTok{                      }\KeywordTok{ends_with}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(year)))\}}
\NormalTok{  temp }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{pivot_longer}\NormalTok{(}\OperatorTok{!}\NormalTok{CASEID_}\DecValTok{1979}\NormalTok{,}
                 \DataTypeTok{names_to =} \StringTok{"job"}\NormalTok{,}
                 \DataTypeTok{values_to =} \StringTok{"hours_work"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{separate}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\KeywordTok{c}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\StringTok{"year"}\NormalTok{), }\DataTypeTok{sep =} \DecValTok{-4}\NormalTok{) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{job =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"job_"}\NormalTok{, }\KeywordTok{substr}\NormalTok{(job, }\DecValTok{9}\NormalTok{, }\DecValTok{10}\NormalTok{))) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ CASEID_}\DecValTok{1979}\NormalTok{)}
\NormalTok{\}}

\CommentTok{# getting the hours of work of all observations}
\NormalTok{years <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\DecValTok{1979}\OperatorTok{:}\DecValTok{1994}\NormalTok{, }\KeywordTok{seq}\NormalTok{(}\DecValTok{1996}\NormalTok{, }\DecValTok{2018}\NormalTok{, }\DataTypeTok{by =} \DecValTok{2}\NormalTok{))}
\NormalTok{hours_all <-}\StringTok{ }\KeywordTok{map_dfr}\NormalTok{(years, get_hour)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{get_rate <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(year) \{}
\NormalTok{  new_data_qnames }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{select}\NormalTok{(CASEID_}\DecValTok{1979}\NormalTok{,}
                  \KeywordTok{starts_with}\NormalTok{(}\StringTok{"HRP"}\NormalTok{) }\OperatorTok{&}
\StringTok{                    }\KeywordTok{ends_with}\NormalTok{(}\KeywordTok{as.character}\NormalTok{(year))) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{pivot_longer}\NormalTok{(}\OperatorTok{!}\NormalTok{CASEID_}\DecValTok{1979}\NormalTok{, }\DataTypeTok{names_to =} \StringTok{"job"}\NormalTok{, }\DataTypeTok{values_to =} \StringTok{"rate_per_hour"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{separate}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\KeywordTok{c}\NormalTok{(}\StringTok{"job"}\NormalTok{, }\StringTok{"year"}\NormalTok{), }\DataTypeTok{sep =} \DecValTok{-4}\NormalTok{) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{job =} \KeywordTok{paste0}\NormalTok{(}\StringTok{"job_0"}\NormalTok{, }\KeywordTok{substr}\NormalTok{(job, }\DecValTok{4}\NormalTok{, }\DecValTok{4}\NormalTok{))) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{id =}\NormalTok{ CASEID_}\DecValTok{1979}\NormalTok{)}
\NormalTok{\}}
\NormalTok{rates_all <-}\StringTok{ }\KeywordTok{map_dfr}\NormalTok{(years, get_rate)}
\CommentTok{# join hours and rates variable}
\NormalTok{hours_wages <-}\StringTok{ }\KeywordTok{left_join}\NormalTok{(rates_all,}
\NormalTok{                         hours_all,}
                         \DataTypeTok{by =} \KeywordTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"year"}\NormalTok{, }\StringTok{"job"}\NormalTok{)) }\OperatorTok{%>%}
\StringTok{  }\CommentTok{# set the 0 value in rate_per_hour as NA}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{rate_per_hour =} \KeywordTok{ifelse}\NormalTok{(rate_per_hour }\OperatorTok{==}\StringTok{ }\DecValTok{0}\NormalTok{, }\OtherTok{NA}\NormalTok{,}
\NormalTok{                                rate_per_hour),}
         \DataTypeTok{hours_work =} \KeywordTok{ifelse}\NormalTok{(hours_work }\OperatorTok{==}\StringTok{ }\DecValTok{0} \OperatorTok{|}\StringTok{ }\NormalTok{hours_work }\OperatorTok{>}\StringTok{ }\DecValTok{84}\NormalTok{, }\OtherTok{NA}\NormalTok{, hours_work))}
\end{Highlighting}
\end{Shaded}

@@

{\centering \includegraphics[width=0.93\linewidth]{/Users/deeamaliah/Documents/summer_project/summer-wages-refresh/paper/figures/flowchart}

@@

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``Robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
Loo, Mark van der. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-rsvg}{}%
Ooms, Jeroen. 2020. \emph{Rsvg: Render Svg Images into Pdf, Png, Postscript, or Bitmap Arrays}. \url{https://CRAN.R-project.org/package=rsvg}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.

@@

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-validate}{}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
